---
title: "NBA Finals Lineup Tracker"
author: "Lindsey Hornberger"
date: "2026-06-23"
output: html_document
---

```{r}
library(hoopR)
library(tidyverse)
library(ggplot2)
```

```{r}
all_pbp <- load_nba_pbp(seasons = 2026)

pbp_g1 <- all_pbp |> filter(game_id == 401859963)
pbp_g2 <- all_pbp |> filter(game_id == 401859964)
pbp_g3 <- all_pbp |> filter(game_id == 401859965)
pbp_g4 <- all_pbp |> filter(game_id == 401859966)
pbp_g5 <- all_pbp |> filter(game_id == 401859967)
```

```{r}
# Function to process one game
process_game <- function(pbp, game_num, spurs_id, knicks_id, 
                         spurs_starters, knicks_starters) {
  
  subs <- pbp |>
    filter(type_text == "Substitution") |>
    select(period_number, start_game_seconds_remaining,
           athlete_id_1, athlete_id_2, team_id, text) |>
    arrange(desc(start_game_seconds_remaining))
  
  spurs <- track_lineups(subs, spurs_id, spurs_starters) |>
    mutate(team = "Spurs", game = game_num)
  
  knicks <- track_lineups(subs, knicks_id, knicks_starters) |>
    mutate(team = "Knicks", game = game_num)
  
  bind_rows(spurs, knicks)
}

# Define star players
spurs_stars <- c(4066259, 5104157, 4845367)  # Fox, Wembanyama, Castle
knicks_stars <- c(3934672, 3136195, 3934719)  # Brunson, Towns, Anunoby

# Detailed star categorization
categorize_lineup_detailed <- function(lineup_str, stars, star_names) {
  ids <- as.integer(str_split(lineup_str, "-")[[1]])
  on_court <- star_names[stars %in% ids]
  n <- length(on_court)
  
  if (n == 3) return("All 3")
  if (n == 2) return(paste(sort(on_court), collapse = " + "))
  if (n == 1) return(on_court)
  return("None")
}

spurs_star_ids   <- c(4066259, 5104157, 4845367)
spurs_star_names <- c("Fox", "Wemby", "Castle")

knicks_star_ids   <- c(3934672, 3136195, 3934719)
knicks_star_names <- c("Brunson", "Towns", "OG")

all_games <- all_games |>
  mutate(
    star_group = case_when(
      team == "Spurs"  ~ map_chr(lineup, ~categorize_lineup_detailed(
                                   .x, spurs_star_ids, spurs_star_names)),
      team == "Knicks" ~ map_chr(lineup, ~categorize_lineup_detailed(
                                   .x, knicks_star_ids, knicks_star_names))
    )
  )

all_games |>
  group_by(team, star_group) |>
  summarise(total_seconds = sum(duration), .groups = "drop") |>
  arrange(team, desc(total_seconds))
```

# build colors palette 
```{r}
# Color palette
# Spurs
spurs_colors <- c(
  "All 3"          = "#1A1A1A",  
  "Fox + Wemby"    = "#4A4A4A",  
  "Castle + Wemby" = "#7A7A7A", 
  "Castle + Fox"   = "#A8A8A8",  
  "Fox"            = "#C8102E",  
  "Wemby"          = "#E8604C", 
  "Castle"         = "#F4A582", 
  "None"           = "#E0E0E0"   
)

# Knicks
knicks_colors <- c(
  "All 3"          = "#003DA5",  
  "Brunson + OG"   = "#0060C0",  
  "OG + Towns"     = "#00A86B",  
  "Brunson + Towns"= "#90BBE8",  
  "Brunson"        = "#F58426",  
  "Towns"          = "#8B2FC9", 
  "OG"             = "#FAC99A", 
  "None"           = "#E0E0E0"
)

# Prep plot data
plot_data <- all_games |>
  mutate(
    x_start    = 2880 - time_start,
    x_end      = 2880 - time_end,
    game_label = factor(paste("Game", game), 
                        levels = paste("Game", 5:1))
  )

# Build chart function
make_chart <- function(data, team_name, color_map, subtitle_text) {
  data |>
    filter(team == team_name) |>
    mutate(star_group = factor(star_group, levels = names(color_map))) |>
    ggplot(aes(xmin = x_start, xmax = x_end,
               ymin = as.numeric(game_label) - 0.4,
               ymax = as.numeric(game_label) + 0.4,
               fill = star_group)) +
    geom_rect(color = NA, linewidth = 0.3) +
    geom_vline(xintercept = c(720, 1440, 2160),
               linetype = "dashed", color = "gray50", linewidth = 0.5) +
    scale_fill_manual(values = color_map, drop = FALSE) +
    scale_x_continuous(
        breaks = c(360, 1080, 1800, 2520),
        labels = c("Q1", "Q2", "Q3", "Q4"),
        limits = c(0, 2880),
        expand = c(0, 0)  
      ) +
    scale_y_continuous(
      breaks = 1:5,
      labels = levels(data |> 
                        filter(team == team_name) |> 
                        pull(game_label))
    ) +
    labs(
      title    = paste(team_name, "— Star Lineup Tracker | 2026 NBA Finals"),
      subtitle = subtitle_text,
      x = NULL, y = NULL,
      fill = NULL
    ) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      axis.text.x     = element_text(size = 10, face = "bold"),
      axis.text.y     = element_text(size = 10, face = "bold"),
      legend.position = "bottom",
      legend.text     = element_text(size = 9),
      plot.title      = element_text(face = "bold", size = 13),
      plot.subtitle   = element_text(size = 10, color = "gray40")
    ) +
    guides(fill = guide_legend(nrow = 2))
}

p_spurs <- make_chart(plot_data, "Spurs",
                      spurs_colors,
                      "Playing time of Fox + Wembanyama + Castle during series")

p_knicks <- make_chart(plot_data, "Knicks",
                       knicks_colors,
                       "Playing time of Brunson + Towns + Anunoby during series")

p_spurs
p_knicks
```

```{r}
# Calculate percentage breakdown per team per game
pct_data <- all_games |>
  group_by(team, game, star_group) |>
  summarise(seconds = sum(duration), .groups = "drop") |>
  group_by(team, game) |>
  mutate(pct = seconds / sum(seconds) * 100) |>
  ungroup() |>
  mutate(
    game_label = paste("Game", game),
    star_group = factor(star_group, levels = c(
      "All 3", 
      "Fox + Wemby", "Castle + Wemby", "Castle + Fox",
      "Fox", "Wemby", "Castle",
      "Brunson + OG", "OG + Towns", "Brunson + Towns",
      "Brunson", "Towns", "OG",
      "None"
    ))
  )

# Lollipop chart function
make_lollipop <- function(data, team_name, color_map, subtitle_text) {
  data |>
    filter(team == team_name) |>
    mutate(star_group = factor(star_group, levels = names(color_map))) |>
    ggplot(aes(x = pct, y = star_group, color = star_group)) +
    geom_segment(aes(x = 0, xend = pct, 
                     y = star_group, yend = star_group),
                 linewidth = 0.8) +
    geom_point(size = 4) +
    geom_text(aes(label = paste0(round(pct, 1), "%")),
              hjust = -0.3, size = 3, color = "gray30") +
    facet_wrap(~game_label, nrow = 1) +
    scale_color_manual(values = color_map, drop = FALSE) +
    scale_x_continuous(limits = c(0, 75),
                       breaks = c(0, 25, 50),
                       labels = c("0%", "25%", "50%")) +
    labs(
      title    = paste(team_name, "— Star Group Time Share by Game | 2026 NBA Finals"),
      subtitle = subtitle_text,
      x = "% of game time", y = NULL
    ) +
    theme_bw() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      strip.text         = element_text(face = "bold", size = 10),
      axis.text.y        = element_text(size = 9),
      axis.text.x        = element_text(size = 8),
      legend.position    = "none",
      plot.title         = element_text(face = "bold", size = 13),
      plot.subtitle      = element_text(size = 10, color = "gray40")
    )
}

lollipop_spurs <- make_lollipop(
  pct_data, "Spurs", spurs_colors,
  "% of game time with Fox + Wembanyama + Castle on floor"
)

lollipop_knicks <- make_lollipop(
  pct_data, "Knicks", knicks_colors,
  "% of game time with Brunson + Towns + Anunoby on floor"
)

lollipop_spurs
lollipop_knicks
```











```{r}
# Get score differential at every pbp moment for all 5 games
score_data <- bind_rows(
  pbp_g1 |> mutate(game = 1),
  pbp_g2 |> mutate(game = 2),
  pbp_g3 |> mutate(game = 3),
  pbp_g4 |> mutate(game = 4),
  pbp_g5 |> mutate(game = 5)
) |>
  select(game, start_game_seconds_remaining, 
         home_score, away_score,
         home_team_id, away_team_id) |>
  mutate(
    spurs_margin = case_when(
      home_team_id == 24 ~ home_score - away_score,
      away_team_id == 24 ~ away_score - home_score
    )
  ) |>
  arrange(game, desc(start_game_seconds_remaining))

# Convert lineup segments to per-second lookup for each team
expand_segments <- function(segs, team_name) {
  segs |>
    filter(team == team_name) |>
    filter(floor(time_start) >= ceiling(time_end)) |>  # drop rounding conflicts
    mutate(second = map2(time_end, time_start, 
                         ~seq(ceiling(.x), floor(.y), by = 1))) |>
    unnest(second) |>
    select(game, second, lineup, star_group)
}

spurs_by_second <- expand_segments(all_games, "Spurs")
knicks_by_second <- expand_segments(all_games, "Knicks")

# Join both teams by game and second
# Join both teams by game and second
matchup_data <- spurs_by_second |>
  inner_join(knicks_by_second,
             by = c("game", "second"),
             suffix = c("_spurs", "_knicks")) |>
  rename(
    spurs_lineup = lineup_spurs,
    knicks_lineup = lineup_knicks,
    spurs_stars  = star_group_spurs,
    knicks_stars = star_group_knicks
  )

# Join score data
score_lookup <- score_data |>
  mutate(second = round(start_game_seconds_remaining)) |>
  group_by(game, second) |>
  slice_tail(n = 1) |>
  ungroup() |>
  select(game, second, spurs_margin)

matchup_data <- matchup_data |>
  left_join(score_lookup, by = c("game", "second"))

matchup_data <- matchup_data |>
  arrange(game, desc(second)) |>
  group_by(game) |>
  fill(spurs_margin, .direction = "down") |>
  ungroup()

# Now calculate point differential per second for each matchup
# We need margin change, not raw margin
matchup_data <- matchup_data |>
  arrange(game, desc(second)) |>
  group_by(game) |>
  mutate(margin_change = lag(spurs_margin) - spurs_margin) |>
  ungroup()

# Aggregate by Spurs lineup vs Knicks star group
matchup_summary <- matchup_data |>
  group_by(spurs_stars, knicks_stars) |>
  summarise(
    seconds = n(),
    total_margin = sum(margin_change, na.rm = TRUE),
    net_per_48 = (total_margin / seconds) * 2880,
    .groups = "drop"
  ) |>
  filter(seconds >= 60) |>  # minimum 1 minute together
  arrange(knicks_stars, desc(net_per_48))

matchup_summary
```


```{r}
# Filter to meaningful matchups only
heat_data <- matchup_summary |>
  filter(seconds >= 60) |>
  mutate(
    spurs_stars = factor(spurs_stars, levels = c(
      "All 3", "Fox + Wemby", "Castle + Wemby", "Castle + Fox",
      "Fox", "Wemby", "Castle"
    )),
    knicks_stars = factor(knicks_stars, levels = c(
      "All 3", "Brunson + OG", "OG + Towns", "Brunson + Towns",
      "Brunson", "Towns", "OG", "None"
    )),
    # Cap extreme values for readability
    net_capped = pmax(pmin(net_per_48, 50), -50)
  )

# Add all combinations including missing ones
all_combos <- expand.grid(
  spurs_stars = levels(heat_data$spurs_stars),
  knicks_stars = levels(heat_data$knicks_stars),
  stringsAsFactors = FALSE
)

heat_data_full <- all_combos |>
  left_join(heat_data, by = c("spurs_stars", "knicks_stars")) |>
  mutate(
    spurs_stars = factor(spurs_stars, levels = levels(heat_data$spurs_stars)),
    knicks_stars = factor(knicks_stars, levels = levels(heat_data$knicks_stars)),
    met_threshold = !is.na(net_per_48),
    label = case_when(
      !met_threshold ~ "",
      TRUE ~ as.character(round(net_per_48, 1))
    )
    
  )|>filter(knicks_stars != "None")

ggplot(heat_data_full, aes(x = knicks_stars, y = spurs_stars)) +
  geom_tile(data = filter(heat_data_full, !met_threshold),
            fill = "#D3D3D3", color = "white", linewidth = 0.5) +
  geom_tile(data = filter(heat_data_full, met_threshold),
            aes(fill = net_capped), color = "white", linewidth = 0.5) +
  geom_text(aes(label = label),
            size = 3.5, fontface = "bold",
            color = ifelse(abs(heat_data_full$net_capped) > 25 & 
                          heat_data_full$met_threshold, "white", "gray20")) +
  scale_fill_gradient2(
    low      = "#C8102E",
    mid      = "#FFF8E7",   # cream
    high     = "#1D428A",
    midpoint = 0,
    limits   = c(-50, 50),
    name     = "Net Pts/48\n(Spurs perspective)",
    na.value = "#D3D3D3"
  ) +
  labs(
    title    = "Spurs Matchup Matrix — 2026 NBA Finals",
    subtitle = "Net points per 48 min | Aggregated across Games 1–5",
caption  = "Blue = good for Spurs · Red = good for Knicks · Gray = insufficient sample (<60 sec)",
    x        = "Knicks Star Unit on Floor",
    y        = "Spurs Star Unit on Floor"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y     = element_text(size = 9),
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, color = "gray40"),
    legend.position = "right"
  )
```
Insight: Fox + Wemby was the Spurs' most effective counter against nearly every Knicks configuration, but the Spurs consistently under-deployed it in favor of their full star unit which repeatedly got outplayed.


more selective matrix
```{r}
heat_data_selective <- all_combos |>
  left_join(heat_data, by = c("spurs_stars", "knicks_stars")) |>
  filter(
    !knicks_stars %in% c("OG", "Towns", "None"),
    !spurs_stars %in% c("Wemby", "Castle + Fox", "Fox"),
    !is.na(spurs_stars), !is.na(knicks_stars)
  ) |>
  mutate(
    spurs_stars = factor(as.character(spurs_stars), levels = c(
      "All 3", "Fox + Wemby", "Castle + Wemby", "Castle"
    )),
    knicks_stars = factor(as.character(knicks_stars), levels = c(
      "All 3", "Brunson + OG", "OG + Towns", "Brunson + Towns", "Brunson"
    )),
    met_threshold = !is.na(net_per_48),
    label = case_when(
      !met_threshold ~ "",
      TRUE ~ as.character(round(net_per_48, 1))
    )
  ) |>
  droplevels()
  

ggplot(heat_data_selective, aes(x = knicks_stars, y = spurs_stars)) +
  geom_tile(data = filter(heat_data_selective, !met_threshold),
            fill = "#989898", color = "white", linewidth = 0.5) +
  geom_tile(data = filter(heat_data_selective, met_threshold),
            aes(fill = net_capped), color = "white", linewidth = 0.5) +
  geom_text(aes(label = label),
            size = 3.5, fontface = "bold",
            color = ifelse(abs(heat_data_selective$net_capped) > 25 & 
                          heat_data_selective$met_threshold, "white", "gray20")) +
  scale_y_discrete(limits = c("All 3", "Fox + Wemby", "Castle + Wemby", "Castle")) +
  scale_x_discrete(limits = c("All 3", "Brunson + OG", "OG + Towns", "Brunson + Towns", "Brunson")) +
  scale_fill_gradient2(
    low      = "#C8102E",
    mid      = "#FFF8E7",   # cream
    high     = "#1D428A",
    midpoint = 0,
    limits   = c(-50, 50),
    name     = "Net Pts/48\n(Spurs perspective)",
    na.value = "#989898"
  ) +
  labs(
    title    = "Spurs Matchup Matrix — 2026 NBA Finals",
    subtitle = "Net points per 48 min | Aggregated across Games 1–5 \nBlue = Spurs | Red = Knicks | Gray = insufficient sample",
caption  = "Fox + Wemby was the Spurs' most effective counter across nearly every Knicks configuration,\nyet the Spurs consistently over-deployed their full star unit — which got outplayed in the same matchups.",
    x        = "Knicks Star Unit on Floor",
    y        = "Spurs Star Unit on Floor"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1, size = 12),
    axis.text.y     = element_text(size = 12),
    axis.title  = element_text(face = "bold"),
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, color = "gray40"),
    plot.caption = element_text(size = 10, color = "black", hjust = 0.5),
    legend.position = "right")
```

Fox + Wemby was the Spurs' most effective counter against nearly every Knicks configuration, but the Spurs consistently under-deployed it in favor of their full star unit which repeatedly got outplayed.


